### 2015-05-27 technical-background.tex
Symmetric cipher theory

\subsection{Symmetric cryptography}
Talk about encryption, integrity and authentication.

Look in depth into AES, NIST approved five modes~\cite{nist-sp800-38A}: CBC (Cipher Block Chaining), ECB (Electronic CodeBook), CFB (Cipher FeedBack), OFB (Output FeedBack) and CTR (Counter).

\begin{figure}
\includegraphics[width=\textwidth]{nist-cbc}
\caption{CBC encryption and decryption diagram}{taken from the NIST recommendation~\cite{nist-sp800-38A}.}
\label{fig:cbc-encrypt-decrypt}
\end{figure}

AEAD => Encryption and authentication: GCM.

>>>>

\subsection{Symmetric cryptography}
Consider a cryptosystem $(P,C,K,E,D)$, with $P$ and $C$ the spaces of plaintexts and ciphertexts, $E$ and $D$ the encryption and decryption algorithms and a keyspace $K$.
\citet{Menezes1996} and \citet{infof405} define a symmetric scheme as a cryptosystem with two keys $e \in K$ and $d \in K$ respectively associated with the encryption and decryption algorithms.
When using the key pair ($e$, $d$), it should be computationally easy to compute $d$ knowing only $e$, and $e$ from $d$.
In the rest of this work, we will consider the most common case: $e = d$, having thus a single key shared between the peers.

There exists numerous cipher types, and one of the most widespread is the block cipher.
A block cipher splits the message into blocks of fixed size and work on them separately.
An example of such a cipher is illustrated on the figure~\ref{fig:cbc-encrypt} with the Cipher Block Chaining (CBC) mode.
This particular mode is easy to implement and has been robust for a long time, before the rise of the padding oracle attack~\cite{vaudenay2002}.

\indent An other interresting block cipher mode is the Counter (CTR).
Each block of data is treated independently from the others (see figure~\ref{fig:ctr-encrypt}, making the scheme parallelizable and thus more interresting on hardware.

\begin{figure}
\includegraphics[width=\textwidth]{nist-cbc}
\caption{CBC encryption diagram}{The output of one block is used as the input of the next one. The decryption is very similar. Taken from the NIST recommendation~\cite{nist-sp800-38A}.}
\label{fig:cbc-encrypt}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{nist-ctr}
\caption{CTR encryption diagram}{Each block is processed independently from the others. The decryption is very similar. Taken from the NIST recommendation~\cite{nist-sp800-38A}.}
\label{fig:ctr-encrypt}
\end{figure}

Those modes can not work alone and have to be associated with a cipher (``CIPH" on figures~\ref{fig:cbc-encrypt}~and~\ref{fig:ctr-encrypt}).
In 1997, the NIST hosted a competition to choose the successor of the Data Encryption Standard (DES).
In 2000, Rijndael emerged victorious and was renamed Advanced Encryption Standard (AES).
It is still nowadays the most used symmetric block cipher used in a large variety of applications.\newline{}

Most applications want to use confidential and authenticated channels for their sensible communication.
In order to do so, they need to combine a symmetric cipher and a MAC.

\noindent New block cipher have emerged to answer the need for such combinations and formed the Authenticated Encryption with Associated Data (AEAD) mode of operation.
The most widespread mode in network applications is the Galois Counter Mode (GCM, see figure~\ref{fig:gcm-encrypt}), combining the CTR block cipher mode and a GHASH function using multiplications in the Galois field.
The advantage of this mode is that the authentication can be computed in parallel with the encryption, itself parallelizable.
This mode can thus be highly optimized in hardware, as proved Barco Silex with their BA415 IP core reaching up to 100Gbps~\cite{barco-ba415}.
